Mendelian randomization for causal inference in Genome-wide association studies

### 2022-02-12 mr-review.Rmd
text about CAUSE

@@

In the short time since publication of @verbanck_detection_2018, researchers have recognized two types of horizontal pleiotropy [@yuan_likelihood_2021]. The first occurs via 
exposure-independent paths. The resulting horizontal pleiotropic effects are independent of the
SNP-exposure relationships. The second type of horizontal pleiotropy manifests when 
in the presence of unobserved exposure-outcome confounding. It induces correlation between 
horizontal pleiotropic effects and SNP-exposure effects. Both types of horizontal pleiotropy 
violate standard MR modeling assumptions and can bias causal effect estimates and can increase false 
discoveries [@yuan_likelihood_2021].
Early MR analyses avoided confounding from horizontal pleiotropy by discarding 
instrumental SNPs that might be associated with the outcome 
[@verbanck_detection_2018;@zhu_causal_2018;@zeng_causal_2019-1;@zeng_causal_2019-2].
More recent methods have attempted to model horizontal pleiotropy 
[@morrison_mendelian_2020;@yuan_testing_2020;@yuan_likelihood_2021]. CAUSE [@morrison_mendelian_2020] and 
MRMix [@qi_mendelian_2019] both use a mixture of normal distributions to control for both 
types of horizontal pleiotropy. Modeling both types of horizontal pleiotropy is particularly 
challenging because the MR model likelihood often involves an integral that can't be solved analytically. 
Because of this issue with the model likelihood, both MRMix and CAUSE use other, non-likelihood-based methods for inference.
MRMix uses ...
CAUSE compares the out-of-sample prediction accuracies for two competing models that differ in inclusion of 
the causal effect parameter [@morrison_mendelian_2020]. Specifically, CAUSE calculates the expected pointwise log posterior density 
between the two models. 
MRAID models horizontal pleiotropy by ...
@yuan_likelihood_2021 found that MRMix is not robust to misspecification of SNP effect sizes and often is 
biased. CAUSE, they found, yields overly conservative p-values [@yuan_likelihood_2021].

@@

The second type of horizontal pleiotropy, from above, is sometimes called "correlated horizontal pleiotropy". Correlated horizontal pleiotropy occurs when 
CAUSE and MRAID are two methods that aim to model and to account for correlated horizontal pleiotropy.

>>>>

and $\tau$ is a vector of independent and identically distributed normal random errors with mean zero.

@@

In the short time since publication of @verbanck_detection_2018, researchers have recognized two 
types of horizontal pleiotropy [@yuan_likelihood_2021]. The first occurs via 
exposure-independent paths. The resulting horizontal pleiotropic effects are independent of the
SNP-exposure relationships. The second type of horizontal pleiotropy manifests when 
in the presence of unobserved exposure-outcome confounding. It induces correlation between 
horizontal pleiotropic effects and SNP-exposure effects. Both types of horizontal pleiotropy 
violate standard MR modeling assumptions and can bias causal effect estimates and can increase false 
discoveries [@yuan_likelihood_2021].
Early MR analyses avoided confounding from horizontal pleiotropy by discarding 
instrumental SNPs that might be associated with the outcome 
[@verbanck_detection_2018;@zhu_causal_2018;@zeng_causal_2019-1;@zeng_causal_2019-2].
More recent methods have attempted to model horizontal pleiotropy 
[@morrison_mendelian_2020;@yuan_testing_2020;@yuan_likelihood_2021]. CAUSE [@morrison_mendelian_2020] and 
MRMix [@qi_mendelian_2019] both use a mixture of normal distributions to control for both 
types of horizontal pleiotropy. Modeling both types of horizontal pleiotropy is particularly 
challenging because the MR model likelihood often involves an integral that can't be solved analytically. 
Because of this issue with the model likelihood, both MRMix and CAUSE use other, non-likelihood-based methods for inference.
CAUSE compares the out-of-sample prediction accuracies for two competing models that differ in inclusion of 
the causal effect parameter [@morrison_mendelian_2020]. Specifically, CAUSE calculates the expected pointwise log posterior density 
between the two models. 
@yuan_likelihood_2021 found that MRMix is not robust to misspecification of SNP effect sizes and often is 
biased. CAUSE, they found, yields overly conservative p-values [@yuan_likelihood_2021].

@@

The second type of horizontal pleiotropy, from above, is sometimes called "correlated horizontal pleiotropy". Correlated horizontal pleiotropy occurs when 
CAUSE and MRAID are two methods that aim to model and to account for correlated horizontal pleiotropy. 
Their approaches to modeling correlated horizontal pleiotropy differ in important ways that, ultimately, 
affect performance. 
CAUSE allows for a proportion of variants to exhibit correlated pleiotropy, which @morrison_mendelian_2020 
models as an effect on a shared, unobserved factor. The remaining variants are independent of this 
unobserved factor. Every variant can have a nonzero pleiotropic effect on the outcome, and these 
pleiotropic effects are uncorrelated with the variant effects on the exposure. The model, then, 
is written as a mixture of biavariate normal distributions, and inference proceeds by borrowing ideas on 
adaptive shrinkage from @stephens_false_2016. Finally, @morrison_mendelian_2020 compares two models 
to determine whether the GWAS summary statistics are consistent with a causal effect of the
exposure on the outcome. Specifically, they estimate the difference in the expected log pointwise 
posterior density for the model with causal effect fixed at zero and the model that permits a nonzero 
causal effect [@vehtari_practical_2017].
